---
title: "Midterm_EDA"
author: "Wenshuo Cai"
date: "2023-11-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(scales)
library(dplyr)
library(lubridate)
library(ggplot2)
library(stringr)
```

## Data Acquisition and Assessment

```{r}
#| echo: false
# Load the data
flood_data <- read.csv("DisasterDeclarationsSummaries.csv")
fema_data<-read.csv("FemaWebDisasterSummaries.csv")
# Convert date columns to Date objects
flood_data$declarationDate <- as.Date(lubridate::ymd_hms(flood_data$declarationDate))
flood_data$incidentBeginDate <- as.Date(lubridate::ymd_hms(flood_data$incidentBeginDate))
flood_data$incidentEndDate <- as.Date(lubridate::ymd_hms(flood_data$incidentEndDate))

# Filter the data for floods in 2020-2021
flood_data_filtered <- flood_data %>%
  filter(incidentType == "Flood", declarationDate >= as.Date('2020-01-01') & declarationDate <= as.Date('2021-12-31'))

# Check for missing values
summary(is.na(flood_data_filtered))

# Remove duplicates
flood_data_filtered <- distinct(flood_data_filtered)


```
## Data cleaning

```{r}


# Removing unneeded columns
flood_data_filtered <- flood_data_filtered %>%
  select(-c(femaDeclarationString, declarationType, tribalRequest))

# Extracting County and Reservation information
flood_data_filtered <- flood_data_filtered %>%
  mutate(CountyArea = ifelse(str_detect(designatedArea, "County"), str_remove(designatedArea, "\\(County\\)"), NA),
         ReservationArea = ifelse(str_detect(designatedArea, "Indian Reservation"), designatedArea, NA)) %>%
  select(-designatedArea)

# Moving new columns to the appropriate place
flood_data_filtered <- flood_data_filtered %>%
  relocate(CountyArea, .after = state) %>%
  relocate(ReservationArea, .after = CountyArea)

# Creating year and month for begin and end dates
flood_data_filtered <- flood_data_filtered %>%
  mutate(BeginYear = year(incidentBeginDate),
         BeginMonth = month(incidentBeginDate),
         EndYear = year(incidentEndDate),
         EndMonth = month(incidentEndDate))

# Calculating the flood duration in days
flood_data_filtered <- flood_data_filtered %>%
  mutate(FloodDuration = as.numeric(incidentEndDate - incidentBeginDate) + 1)

# Converting to numeric types for year and month columns if needed
flood_data_filtered$BeginYear <- as.numeric(flood_data_filtered$BeginYear)
flood_data_filtered$BeginMonth <- as.numeric(flood_data_filtered$BeginMonth)
flood_data_filtered$EndYear <- as.numeric(flood_data_filtered$EndYear)
flood_data_filtered$EndMonth <- as.numeric(flood_data_filtered$EndMonth)



# Print the column names of the data frame
#print(colnames(fema_data))

# Clean fema_data
fema_data <- fema_data |>
    semi_join(flood_data_filtered, by = "disasterNumber")
fema_data <- fema_data |>
    select(disasterNumber, totalNumberIaApproved, totalAmountIhpApproved, totalAmountHaApproved, totalAmountOnaApproved, totalObligatedAmountHmgp)

flood_data_cleaned<- flood_data_filtered |>
    left_join(fema_data, by = "disasterNumber")

# Define the state dictionary as a named vector
state_dict <- c(
  WA = "Washington",
  VA = "Virginia",
  NH = "New Hampshire",
  VT = "Vermont",
  AZ = "Arizona",
  TN = "Tennessee",
  LA = "Louisiana",
  WV = "West Virginia",
  HI = "Hawaii",
  KY = "Kentucky",
  PR = "Puerto Rico",
  ND = "North Dakota",
  OR = "Oregon",
  WI = "Wisconsin"
)

# Replace state abbreviations with full names in the 'state' column of the 'flood' data frame
flood_data_cleaned$state <- sapply(flood_data_cleaned$state, function(abbrev) {
  if (abbrev %in% names(state_dict)) {
    return(state_dict[[abbrev]])
  } else {
    return(abbrev)
  }
})

# Remove "(County)" from the 'County' column and trim whitespace
flood_data_cleaned$County <- gsub(" \\(County\\)", "", flood_data_cleaned$County)
flood_data_cleaned$County <- trimws(flood_data_cleaned$County)

# Remove object 'FA' from the environment
rm(fema_data)

# Export the cleaned data to a new CSV file
write.csv(flood_data_filtered, "flood_data_cleaned.csv", row.names = FALSE)

```


---

## Strom Data assessment

```{r}
storm_d20 <- read.csv("StormEvents_details-ftp_v1.0_d2020_c20230927.csv", header = T)
storm_d21 <- read.csv("StormEvents_details-ftp_v1.0_d2021_c20231017.csv", header = T)
storm_f20 <- read.csv("StormEvents_fatalities-ftp_v1.0_d2020_c20230927.csv",
                           header = T)
storm_f21 <- read.csv("StormEvents_fatalities-ftp_v1.0_d2021_c20231017.csv",
                           header = T)
storm_l20 <- read.csv("StormEvents_locations-ftp_v1.0_d2020_c20230927.csv",
                           header = T)
storm_l21 <- read.csv("StormEvents_locations-ftp_v1.0_d2021_c20231017.csv",
                           header = T)




# Cleaning and selecting relevant columns from the details datasets
Details_2020 <- storm_d20 %>%
    select(ID = EVENT_ID, StartYM = BEGIN_YEARMONTH, StartD = BEGIN_DAY, 
           EndYM = END_YEARMONTH, EndD = END_DAY, Region = STATE, Cause = FLOOD_CAUSE)

Fatalities_2020 <- storm_f20 %>%
    select(ID = EVENT_ID, Age = FATALITY_AGE, Gender = FATALITY_SEX) %>%
    mutate(FatalityCount = 1)

Details_2021 <- storm_d21 %>%
    select(ID = EVENT_ID, StartYM = BEGIN_YEARMONTH, StartD = BEGIN_DAY, 
           EndYM = END_YEARMONTH, EndD = END_DAY, Region = STATE, Cause = FLOOD_CAUSE)

Fatalities_2021 <- storm_f21 %>%
    select(ID = EVENT_ID, Age = FATALITY_AGE, Gender = FATALITY_SEX) %>%
    mutate(FatalityCount = 1)

# Merging details with fatalities data
Combined_2020 <- Details_2020 %>%
    left_join(Fatalities_2020, by = "ID")

Combined_2021 <- Details_2021 %>%
    left_join(Fatalities_2021, by = "ID")

# Create a function to calculate duration
calculate_duration <- function(startYM, startD, endYM, endD) {
    startDate <- as.Date(paste(substr(startYM, 1, 4), substr(startYM, 5, 6), startD, sep = "-"))
    endDate <- as.Date(paste(substr(endYM, 1, 4), substr(endYM, 5, 6), endD, sep = "-"))
    return(as.integer(endDate - startDate) + 1)
}

# Apply the function to calculate the duration of the storm events
Combined_2020 <- Combined_2020 %>%
    mutate(Duration = calculate_duration(StartYM, StartD, EndYM, EndD))

Combined_2021 <- Combined_2021 %>%
    mutate(Duration = calculate_duration(StartYM, StartD, EndYM, EndD))

# Determine if the event is related to flooding
Combined_2020 <- Combined_2020 %>%
    mutate(IsFlood = as.integer(!is.na(Cause) & Cause != ""))

Combined_2021 <- Combined_2021 %>%
    mutate(IsFlood = as.integer(!is.na(Cause) & Cause != ""))

# Organize the Combined_2020 dataset
OrganizedStorm_2020 <- Combined_2020 %>%
  select(
    EventID = ID, 
    State = Region, 
    YearStart = StartYM, 
    MonthStart = StartD, 
    DayStart = StartD, 
    YearEnd = EndYM, 
    MonthEnd = EndD, 
    DayEnd = EndD, 
    EventDuration = Duration, 
    FloodEvent = IsFlood, 
    FloodCause = Cause, 
    FatalityFlag = FatalityCount, 
    FatalityAge = Age, 
    FatalityGender = Gender
  )

# Organize the Combined_2021 dataset
OrganizedStorm_2021 <- Combined_2021 %>%
  select(
    EventID = ID, 
    State = Region, 
    YearStart = StartYM, 
    MonthStart = StartD, 
    DayStart = StartD, 
    YearEnd = EndYM, 
    MonthEnd = EndD, 
    DayEnd = EndD, 
    EventDuration = Duration, 
    FloodEvent = IsFlood, 
    FloodCause = Cause, 
    FatalityFlag = FatalityCount, 
    FatalityAge = Age, 
    FatalityGender = Gender
  )
write.csv(OrganizedStorm_2020, "OrganizedStorm_2020.csv", row.names = FALSE)
write.csv(OrganizedStorm_2021, "OrganizedStorm_2021.csv", row.names = FALSE)
```

---
## EDA

# use plots to understand the impact and patterns of floods
```{r}
flood_data <- read.csv("flood_data_cleaned.csv")
flood_data$declarationDate <- as.Date(flood_data$declarationDate, format="%Y-%m-%d")
flood_data$YearMonth <- format(flood_data$declarationDate, "%Y-%m")
# Aggregate the data by YearMonth
floods_by_month <- flood_data %>%
  group_by(YearMonth) %>%
  summarise(Count = n())

# Plot
ggplot(floods_by_month, aes(x=YearMonth, y=Count)) +
  geom_line(group=1) +
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title="Number of Flood Events by Month and Year", x="Year and Month", y="Number of Flood Events")
```
The line plot shows the number of flood events by month and year. It appears there are fluctuations in the number of flood events over time, with some months experiencing more flood events than others. This kind of pattern can be influenced by seasonal weather patterns, such as heavy rainfall seasons or melting snow.

```{r}
# Plot the distribution of flood durations
ggplot(flood_data, aes(x=FloodDuration)) +
  geom_histogram(bins=30, fill="blue", color="black") +
  labs(title="Distribution of Flood Durations", x="Duration of Flood (Days)", y="Frequency")

```
The histogram illustrates the distribution of flood durations, measured in days. We can observe that the majority of floods last less than 10 days, with a noticeable peak around that duration. There are fewer instances of longer-lasting floods, suggesting that while short-term floods are common, prolonged flooding events are less frequent.

```{r}
# Count the number of flood events by state
floods_by_state <- table(flood_data$state)

# Plot
barplot(floods_by_state, las=2, col="skyblue", main="Number of Flood Events by State", xlab="State", ylab="Number of Flood Events")

```

The bar plot shows the number of flood events by state. It seems that certain states have a higher number of recorded flood events, which may indicate a greater vulnerability to flooding or could be related to the size and geography of the state.

```{r}
# Prepare data for the assistance programs declared
assistance_programs <- c("ihProgramDeclared", "iaProgramDeclared", "paProgramDeclared", "hmProgramDeclared")
assistance_data <- colSums(flood_data[,assistance_programs])

# Sort assistance_data in ascending order for horizontal bar plot
assistance_data_sorted <- sort(assistance_data, decreasing = TRUE)

# Convert to data frame for ggplot
assistance_data_df <- data.frame(Program = names(assistance_data_sorted), Declarations = assistance_data_sorted)

# Plot the declaration of assistance programs
ggplot(assistance_data_df, aes(x = Program, y = Declarations)) +
  geom_bar(stat = "identity", fill = "#008080") +
  coord_flip() +  # Flip coordinates for horizontal bar plot
  labs(title = "Assistance Programs Declared During Floods", x = "Number of Declarations", y = "Assistance Programs") +
  theme_minimal() +
  theme(panel.grid.major.x = element_line(colour = "gray90"))

# Display the plot
ggsave("Assistance_Programs_Declared_During_Floods.png", width = 10, height = 6, dpi = 300)
```

The horizontal bar plot shows the number of times different assistance programs have been declared during floods. The paProgramDeclared (Public Assistance) and hmProgramDeclared (Hazard Mitigation) have been declared the most frequently, suggesting that floods often lead to public infrastructure damage and necessitate long-term risk reduction measures. The iaProgramDeclared (Individual Assistance) has not been declared for any of the events in this dataset, which might indicate that the floods did not meet the threshold for this type of assistance or that the data on individual assistance is not captured.



